---
title: "An introduction to tidyverse"
author: "Jonathan Landry"
date: "`r doc_date()`"
output: 
    BiocStyle::html_document:
        toc: true
        toc_float: true
        highlight: tango
        code_folding: show
    BiocStyle::pdf_document:
        toc: true
        highlight: tango
---


<!--
To compile this document
graphics.off();rm(list=ls());rmarkdown::render('R-lab.Rmd');purl('R-lab.Rmd')
pdf document
rmarkdown::render('R-lab.Rmd', BiocStyle::pdf_document2())
-->

```{r options, include=FALSE}
library(knitr)
options(digits = 3, width = 80)
opts_chunk$set(echo = TRUE, tidy = FALSE, include = TRUE,
               dev = 'png', fig.width = 6, fig.height = 3.5, 
               comment = '  ', dpi = 300,
cache = TRUE)
```

This tutorial has been greatly inspired from Bernd Klaus materials. Many thanks for sharing.


# Required packages and other preparations

```{r required_packages, echo = FALSE, warning=FALSE, message=FALSE, results="hide"}
suppressPackageStartupMessages({
library("TeachingDemos")
library("openxlsx")
library("multtest")
library("Biobase")
library("tidyverse")
library("cowplot")
})
```

```{r required packages and data, echo = TRUE, message=FALSE, warning=FALSE}
library("TeachingDemos")
library("openxlsx")
library("multtest")
library("Biobase")
library("tidyverse")
library("cowplot")
```


# Classes, modes and types of objects
R is an object-oriented language, so every data item is an object in R.
As in other programming languages, objects are instances of "blue-prints" called classes.
There are the following elementary types or ("modes"): 

* numeric: real number
* character: chain of characters, text
* factor: categorical data that takes a fixed set of values
* logical: TRUE, FALSE
* special values:  NA (missing value), NULL ("empty object"),
 Inf,  -Inf (infinity), NaN (not a number) 
 
We haven't met factors yet: They are designed to represent categorical data that 
can take a fixed set of possible values. Factors are built on top of integers, 
and have a levels attribute:

```{r factors}

x <- factor(c("ab", "cd", "ab"), levels = c("ab", "cd", "ef"))
x

```


Data storage types includes matrices, lists, data frames (tibbles), which will be introduced
in the next section. Certain types can have different subtypes, e.g. numeric
can be further subdivided into the integer, single and double types. Types
can be checked by the `is.*` and changed ("casted") by the 
`as.*` functions. Furthermore, the function
`str` is very useful in order to obtain an overview of an (possibly
complex) object at hand. The following examples will make this clear.
We first assign the value `9` to an object and then perform various operations on it. 


```{r object-examples,   echo = TRUE}
a <- 9
# is a a string?
is.character(a) 
# is a a number?
is.numeric(a) 
# What's its type?
typeof(a)
# now turn it into a factor
a <- as.factor(a)
# Is it a factor?
is.factor(a)
# assign an string to a: 
a <- "NAME"
# what's a?
class(a)
str(a) 
```


# Matrices, lists,  data frames and basic data handling

## Matrices

Matrices are two--dimensional vectors and 
can be created in R  in a variety of ways. Perhaps the simplest is to create the columns
and then glue them together with the command `cbind`. For example:


```{r cbind-ex,   echo = TRUE}
x <- c(5, 7 , 9)
y <- c(6, 3 , 4)
z <- cbind(x, y)
z
dim(z)
```

We can also use the function `matrix()` directly to create a matrix.
```{r matrix_direct,   echo = TRUE}
z <- matrix(c(5, 7, 9, 6, 3, 4), nrow = 3)
```

There is a similar command, `rbind`, for building matrices by
gluing rows together.
The functions `cbind` and `rbind` can also be applied to matrices themselves 
(provided the dimensions match) to form larger matrices.

Notice that the dimension of the matrix is  determined
by the size of the vector and the requirement that the number of 
rows is 3 in the example above, as specified by the
argument `nrow = 3`. As an alternative we could have specified the number of columns with the
argument `ncol = 2` (obviously, it is unnecessary to give both). Notice that the matrix is "filled up"
column-wise. If instead you wish to fill up row-wise, add the option `byrow=TRUE`. 
 

```{r Matrix-ex,   echo = TRUE}
z <- matrix(c(5, 7 , 9 , 6 , 3 , 4), nrow = 3, byrow = TRUE)
z
```

R will try to interpret operations on matrices in a natural way. 
For example, with z as above, and y defined below we get:

```{r Matrix-op,   echo = TRUE}
y <- matrix(c(1, 3, 0, 9, 5, -1), nrow = 3, byrow = TRUE)
y
y + z
y * z
```

Notice that multiplication here is component--wise.
As with vectors it is useful to be able to extract sub-components of matrices. In this case, we
may wish to pick out individual elements, rows or columns. As before, the `[ ]` 
notation is used to subscript. The following examples illustrate this:

```{r Matrix-op-4,   echo = TRUE}
z[1, 1]
z[, 2]
z[1:2, ]
z[-1, ]
z[-c(1, 2), ]
```


So, in particular, it is necessary to specify which rows and columns are required, 
whilst omitting the index for either dimension implies that every element 
in that dimension is selected.



## Data frames (tibbles) and lists


A data frame is  a matrix where the columns can have different data types.
As such, it is usually used to represent a whole data set, where the rows
represent the samples and columns the variables. Essentially, you can think
of a data frame as an excel table.

Here, we will meet the first tidyverse member, namely the `r CRANpkg("tibble") ` 
package, which improves the conventional R `data.frame` class. A tibble
is a `data.frame` which a lot of tweaks and more sensible defaults that make your
life easier. For details on the tweaks, see  the help on tibble: `?tibble`
so that you never have to use a standard data frame anymore.

Let's illustrate this by the small data set
saved in comma--separated-format (csv) ---
`patients`. We load it in from a website using the function
`read_csv`, which is used to import a data file in 
*comma separated format --- csv* into R. In a .csv--file the data 
are stored row--wise, and the entries in each row are separated by commas. 

The function `read_csv` is from the `r CRANpkg("readr")` package and will
give us a tibble as the result. The function `glimpse()` gives a nice summary
of a tibble.

It has weight, height and gender of three people. 
We can also use the function `read.xlsx` from the `r CRANpkg("opnxlsx")` 
package to import the  data from an excel sheet. Here, we have to use the
function `as_tibble` to turn the data.frame into an equivalent tibble.

```{r load-Patients-excel,   echo = TRUE}
pat <- as_tibble(read.xlsx("Patients.xlsx"))
pat
glimpse(pat)
str(pat)
```

## Accessing data in data frames

Now that we have imported the small data set, you might be wondering how to 
actually access the data. For this the functions `filter` and `select` from
the `r CRANpkg("dplyr")` package of the `r CRANpkg("tidyverse")` are useful. 
`filter` will select  certain rows (observations),
while `select` will subset the columns (variables of the data). In the following
command, we get all the patients that are less tall than 1.5 and select their Height and
Gender as well as their Id:

```{r subset_data}
pat_tiny <- filter(pat, Height < 1.7)
select(pat_tiny, PatientId,  Height, Gender)
```

There are a couple of operators useful for comparisons:

*  `Variable == value`: equal
*  `Variable != value`: un--equal
*  `Variable < value`: less
*  `Variable > value`: greater
*  `&: and`
*  `|` or
*  `!`: negation
*  `%in%`: is element?

The function `filter` allows us to combine multiple conditions easily, 
if you specify multiple of them, they will automatically concatenated via a `&`.
For example, we can easily get light and female patients via:


```{r light_and_small_patients,   echo = TRUE}
filter(pat, Height < 1.5, Gender == "f")
```
We can also retrieve small OR female patients via

```{r light_or_small_patients,   echo = TRUE}
filter(pat, (Height < 1.5) | (Gender == "f"))
```

## Vectors with arbitrary contents: Lists

Lists can be viewed as vectors that contain not only elementary objects such 
as number or strings but can potentially arbitrary objects. The following example
will make this clear. The list that we create contains a number, two vectors
and a string that is itself part of a list.


```{r list_example,   echo = TRUE}
L <- list(one = 1, two = c(1, 2), five = seq(1, 4, length = 5), 
          list(string = "Hello World"))
L
```

Lists are the most general data type in R. In fact, data frames (tibbles) are
lists with elements of equal lengths. List elements can either be accessed by 
their name using the dollar sign `$` or via their position via a double 
bracket operator `[[]]`.

```{r list_access,   echo = TRUE}
names(L)
L$five + 10
L[[3]] + 10
```

Using only a single bracket (`[]`) will extract a sublist, so the result will
always be a list, while the dollar sign `$` or the double bracket operator
`[[]]` removes a level of the list hierarchy. Thus, in order to access the 
string, we would first have to extract the sublist containing the string from `L`
and then get the actual string from the sublist, `[[` drills down into the list 
while `[` returns a new, smaller list. 

```{r}
L[[4]]$string
L[2]
```

Since data frames are just a special kind of lists, 
they can actually be accessed in the same way.

```{r list-example_df,   echo = TRUE}
pat$Height
pat[[2]]
pat[["Gender"]]
```

More on lists can be found in the respective chapter of "R for data science"
[here](http://r4ds.had.co.nz/vectors.html#lists).

## Applying a function to elements of a data structure

R encourages the use of functions for programming. Instead of e.g. looping through
a vector or data frame, you can use specialized functions that apply another function 
to each element of your data. These kinds of functions are called apply functions. 
Here, we will use the `map` family
of functions from the `r CRANpkg("purrr")` package instead of the base R functions.
An apply / map call applies a function to a vector or list and returns the result in
another vector/list. Thus, each step consists of "mapping" a list value to a result. 

We will introduce the `map` functions  by looking at a typical data set in a tabular
format, where the rows represent the samples and the columns the variables measured. The
data set `bodyfat` contains various body measures for 252 men. We turn
it into a tibble by using the function `as_tibble()`.

Let's inspect it a bit. The first thing we notice
is that tibbles prints only the first 10 rows by default. Tibbles are designed 
so that you don’t accidentally overwhelm your console when you print large data
frames. Additionally, we get a nice summary of the variables available in our
data set.

```{r loadBodyfat,   echo = TRUE}
load("bodyfat.rda")
bodyfat <- as_tibble(bodyfat)
bodyfat
```

As data frames are just a special kind of list, namely a list that is composed of
vectors of equal length, we can use a map function to compute the mean value for every
variable in our data set.

```{r bodyfat_map, dependson = "loadBodyfat"}
head(map_dbl(bodyfat, mean))
```

Here we use map_dbl, to ensure that we get a double value back. There 
are specialized mapping functions for many data types, but you can always
use the default  `map()` function  as a fallback when there is no specialized
equivalent available.

The map functions are really useful for applying your custom functions, for example
we can compute a robust z--score by subtracting the median and dividing by the mean
absolute deviation for each variable. 

This will bring all the variables in the data set to a common scale and make 
them directly comparable. These kinds of transformations are often performed 
before clustering or dimensionality reduction.

You can create your own functions very easily by adhering to the following
template:

```{r function_template, eval=FALSE}
function_name <- function(argument_1, argument_2, 
                          optional_argument = defautl_value ) 
{
  return(...)
}
```

As you can see, the source code of the  function has to be in curly brackets, while
the arguments are defined in the parantheses. Arguments without a default value 
are mandatory, and default value are specified by equality signs.

By default R returns the result of the last computation performed within the 
curly brackets  (often, this will be the last line of the function). However, 
you can always specify the return  value directly with `return()`. If 
you want to return multiple values, you can return a list.

We can now easily define our function and apply it to the data set.

```{r robust_z}
robust_z <- function(x){
  (x - median(x)) / mad(x)
}

head(map_df(bodyfat, robust_z), 3)

```

Here, we used the function `map_df` to make sure that we get a data frame back.
There is an even simpler way to achieve the same goal. Using a tilde (~) to create
an R formula, the map
functions allow you to define anonymous functions with a default argument `.x`.

With this, we do not need to define our robust z--score function explicitly.

```{r robust_z_implicit}
head(map_df(bodyfat, ~  (.x - median(.x)) / mad(.x)), 3)
```

## Computing variables from existing ones and predicate functions

Often, we want to use variables stored in our data set to compute derived quanities.
For example, we might be interest in the weight in kilograms instead of pounds
and the hight in meters instead of inches. The function `mutate` allows us to
do this.

```{r transform_example}
pb_to_kg <- 1/2.2046
inch_to_m <- 0.0254

bodyfat <- mutate(bodyfat, height_m = height * inch_to_m,  
                           weight_kg = weight * pb_to_kg)

select(bodyfat, height,  height_m, weight, weight_kg)
```

We often want to apply our function only to variables in the data set that are 
of a specific type, e.g. numeric, we kann use simple predicate functions that
return `TRUE` or `FALSE` in combination with `discard` or `keep` to perform 
appropriate selections. For example, we can exclude the id column of the patients
data set, before computing the variable--wise means.

```{r predicate_functions}
keep(pat, is_double)
map_dbl(discard(pat, is_character), mean, na.rm = TRUE)
```

Note that we specified `na.rm = TRUE` as an additional argument to the map function.
This will be directly passed on to the argument list of `mean` and excludes any
missing values before computing the means.

### Exercise: Handling a small data set

* Import the data set `Patients.csv` from the website  
* Which variables are stored in the data frame and what are their values?
* Is there a missing weight value?  If yes, replace it by the mean of the other weight values.
* Calculate the mean weight and height of all the patients.  
* Calculate the `BMI = Weight / Height^2`  of all the patients.


# Simple plotting in R: qplot of `r CRANpkg("ggplot2")`

The package `r CRANpkg("ggplot2")` allows very flexible plotting in R, 
but takes a while to get acquainted with the underlying grammer 
of graphics. Thus, we will use its function  `qplot()` for "quick plotting",
which requires no knowledge of the underlying advanced features and behaves
much like R's default `plot` function. 
However, it offers advanced options like facetting or coloring by condition 
as well.


```{r qplot, eval=FALSE}
qplot(x, y = NULL, ..., data, facets = NULL, 
  NA), ylim = c(NA, NA), log = "", main = NULL,
  xlab = , ylab = )

```

The arguments are:

* `x:` x--axis data
* `y:` y--axis data (may be missing)
* `data:`  `data.frame` containing the variables used in the plot 
* `facets ` split the plot into facets, use a formula like
. ~split to do wrapped splitting and row ~ columns to split by rows and columns
* `main:` plot heading
* `color, fill` set to factor/string in the data set in order to
color the plot depending on that factor. Use `I("colorname")` to use a
specific color.
* `geom` specify a "geometry" to be used in the plots, examples
include point, line, boxplot, histogram etc.
* `xlab, ylab, xlim, ylim` set the x--/y--axis parameters


As an example, we create a plot of `perc.fat` against abdomen circumference and
color it by weight. For this we bin the weight vector into 5 discrete categories
using the `cut` function.

```{r qplot_example}
bodyfat <- mutate(bodyfat, weight_binned = cut(weight_kg, 5))

qplot(abdomen.circum, percent.fat, 
      color = weight_binned, data = bodyfat) 

```

We can (unsurprisingly) see that abdomen circumference, weight and bodyfat are highly 
correlated to each other. We can also produce a facetted plot split by weight.

```{r qplot_example_facets}

qplot(abdomen.circum, percent.fat, 
      color = weight_binned, data = bodyfat, 
      facets = ~weight_binned) 

```


### Exercise: Plotting the EMBL logo


The code below plots the embl logo.  The plus sign adds additional 
"layers" to the ggplot object modifying any given plot and we use it here
to make the axes dissapear.

However the colors are not quite right. Can you fix that?
Check out the [ggplot2 docs](http://docs.ggplot2.org/) or try googeling!

```{r embl_logo, results='hide'}

load("hex_grid.Rdata")
embl_colors <- c("#E33E3E", "#73B360")
qplot(x, y, data = hex_grid, color = lab, asp = 1) +
    theme(panel.border = element_blank(), 
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(), 
          axis.text = element_blank(),
          line = element_blank(),
          title = element_blank())+
    scale_color_manual(values=embl_colors)

```

# Programming statements

R offers the typical options for flow--control known from many other
languages. 

The most important one, the **if--statement** is used when certain computations should
only be performed if a certain condition is met (and maybe something else should be performed
when the condition is not met):


```{r if_example,   echo = TRUE, eval = TRUE}
w <- 3
  if (w < 5) {
      d <- 2
    } else {
      d <- 10
    }
d
```


If you want perfom a computation for every entry of a list,
 you usually do  the computations for one time step and then for
the next and the next, etc. Because nobody wants 
to type the same commands over and over again,
these computations are automated in  __for--loops__.

```{r for_example,   echo = TRUE, eval = TRUE}
h <- seq(from = 1, to = 8)
s <- numeric() # create empty  vector
    for (i in 1:8)
    {
    s[i] <- h[i] * 10
    }
s
```

Note however, that you should typically resort to `map` function for this 
purpose as this leads to more readable code:

```{r maps_for}
map_dbl(h, ~.x*10)
```


Another useful command is the __ifelse__--command, it replaces 
elements of a vector based on the evaluation of
another logical vector of the same size. This is useful to replace missing
values, or to binarize a vector.

```{r ifelse-example,   echo = TRUE, eval = TRUE}
s <- seq(from = 1, to = 10)
binary_s <- ifelse(s > 5, "high", "low")
binary_s
```


## Exercise: Base calling errors

The function `readError(noBases)`  simulates the base calling 
  errors of a sequencing machine. The parameter
 \texttt{noBases}  represents the  number of positions in the genome sequenced
 and the function will return a vector, which has the entry "error" if a base 
 calling error occurs at a certain position and "correct" if  the base is 
 read correctly. It can be obtained with the command 

` source("readError.R")`

* Let the sequencer read a thousand positions and try to infer a base calling 
error rate from this simulation
HINT: The functions  `table`  and `prop.table` could be handy for this!
* Let us assume the technology improves and the machine is less prone to  errors.
Change the function accordingly!

# Answers to exercises

__Exercise: Handling a small data set__

* Import the data set `Patients.csv` from local directory 
* Which variables are stored in the data frame and what are their values?
* Is there a missing weight value?  If yes, replace it by the mean of the other weight values.
* Calculate the mean weight and height of all the patients.  
* Calculate the `BMI = Weight / Height^2`  of all the patients.


__Solution: Handling a small data set__

```{r apply-test,   echo = TRUE, results = 'hide', message=FALSE}
pat <- read_csv("Patients.csv")
pat

map_dbl(keep(pat, is_double), mean)

pat$Weight[2] <- mean(pat$Weight, na.rm = TRUE)
pat

pat <- mutate(pat, BMI = Weight / Height^2)
```


__Exercise: Plotting the EMBL logo__

The code below plots the embl logo. Note that the plus sign adds additional 
layers to the ggplot object. This allows you to modify any given plot.
However the colors are not quite right. Can you fix that?
Check out the [ggplot2 docs](http://docs.ggplot2.org/) or try googeling!

```{r embl_logo_ex, results='hide'}

load("hex_grid.Rdata")
embl_colors <- c("#E2001A", "#6FAA46")
qplot(x, y, data = hex_grid, color = lab, asp = 1) +
    theme(panel.border = element_blank(), 
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(), 
          axis.text = element_blank(),
          line = element_blank(),
          title = element_blank())

```

__Solution: Plotting the EMBL logo__

```{r sol_embl_logo, results='hide', fig.keep='none'}

load("hex_grid.Rdata")
embl_colors <- c("#E2001A", "#6FAA46")
 
qplot(x, y, data = hex_grid, color = lab, asp = 1) +
    theme(panel.border = element_blank(), 
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(), 
          axis.text = element_blank(),
          line = element_blank(),
          title = element_blank()) +
    scale_colour_manual(values = embl_colors )

```

__Exercise: Base calling errors__
The function `readError(noBases)`  simulates the base calling 
  errors of a sequencing machine. The parameter
 \texttt{noBases}  represents the  number of positions in the genome sequenced
 and the function will return a vector, which has the entry "error" if a base 
 calling error occurs at a certain position and "correct" if  the base is 
 read correctly. It can be obtained with the command 

` source("readError.R")`

* Let the sequencer read a thousand positions and try to infer a base calling 
error rate from this simulation
HINT: The functions  `table`  and `prop.table` could be handy for this!
* Let us assume the technology improves and the machine is less prone to  errors.
Change the function accordingly!


__Solution: Base calling errors__
```{r sol_read,   echo = TRUE, results = 'hide'}

source("readError.R")

test <- readError(1000)
## number of errors
sum(test == "error")
##  error probability
sum(test == "error") / 1000

prop.table(table(test))


readError2 <- function(noBases){

  positions <- integer(noBases) ## initialize vector
  for (i in 1:noBases ) {	
    positions[i] <- rbinom(n = 1, size = 1, prob = 0.05)
  }
    return(ifelse(positions, "correct", "error"))
  }

### equivalent function
rbinom(n = 1000, size = 1, prob = 0.05)
```

```{r seesionInfo, results='markup'}
sessionInfo()
```
